Two Decades of Smalltalk VM Development

### 2018-08-17 paper.tex
Write the "JIT fidelity and the interfacing of machine code to the simulation objects" paragraph (follows JIT simulation).

The VM executable is generated in a two step process. Firstly, the Slang-to-C compiler translates the Slang code to C code, generating a few files. This first step takes several seconds. Secondly, the C compiler (depending on the platform, LLVM or GCC), translates the C code into an executable. This second step may take several dozens of seconds the first time. But, depending on what the programmer changed and what the C compiler have cached, it usually only takes a few seconds to recompile the C code.

The VM can actually be configured in two main flavours, interpreter-only or interpreter+JIT. Although the version with the JIT is most often used in production, the interpreter version is convenient for development purposes \footnote{and on devices that outlaw JITs}. For example, debugging the garbage collector or evaluating new language features can be done in the interpreter-only VM, avoiding the JIT complexity.

@@

\paragraph{Start-up from snapshots.} 
Smalltalk is an object \emph{system}, rather than a language. The entire system, including its development tools and application code is stored in a snapshot file, which is essentially a memory dump of the entire heap. A Smalltalk system is started by loading a snapshot file into a VM. The snapshot includes objects such as the classes, the compiled methods in the form of bytecodes and the running processes. At start-up, the VM restores the state of all objects in memory and resumes execution in the process active when the snapshot was created.  A new snapshot can be made during or at the end of the session.

When programming with Smalltalk, the programmer usually starts from a snapshot which contains the core libraries, the development environment and the application under development. Development of the application consists essentially in writing and editing code, which effectively installs, modifies and removes classes and compiled methods from the set of live objects. Programming may be done live, as the application under development is running, for example creating or revising methods in the debugger, and continuing.  Objects may have their shape changed on the fly as instance variables are added and removed.
\egb{there is overlap here with respect to the previous paragraph. We already knew that a snapshot can be made during a coding session, and what a snapshot includes}
  When the programmer chooses, he can take a new snapshot, which includes his code changes, and may include application instances under development [MAYBE THIS SHOULD BE DESCRIBED BELOW IN SNAPSHOTS]. \egb{deployment here refers to starting a Smalltalk system from the previous paragraph, right?}Deployment is performed using a snapshot containing the deployed application and the required core libraries (unused libraries and development tools may be removed from the deployed snapshot to decrease memory footprint and for security purposes).  \footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello \cite{Monticello}}.

Starting up \openSmalltalkVM from source files has never been supported. However, recent work~\cite{PharoBootstrap} allows a snapshot to be created from sources, indirectly allowing the VM to be started from source files. 
The bootstrap process uses the simulator to create and manipulate the heap and then to save the resulting snapshot.

\paragraph{Interpreter-only simulation.} 
As the VM can be compiled with and without the JIT, the VM can also be simulated with or without the JIT. Since executing machine code on the processor simulations is not fast, interpreter simulation may be faster, and startup time and memory footprint are smaller.

@@

At start-up, the VM mmaps a memory region used for the executable code generated by the JIT (optional executable section, exists only if the JIT is enabled), for the young objects and for the old objects present in the snapshot started plus a little more space for the first \emph{tenures}\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector. Later during execution, as old space grows, new memory regions are mmap'ed on higher addresses to store other old objects. Usually at very high addresses, we can find the C stack. Room in the C stack is allocated at VM start-up to hold the Smalltalk stack. Both stacks are disjointed and managed differently.

In the simulator, the heap is simulated as a large contiguous byte array. References between objects are effectively indexes inside the byte array instead of pointers. All the Slang variables, normally translated to C variables, are simulated as Smalltalk objects. They used specific classes, such as \emph{CArrayAccessor} over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs, etc.). The Slang code is executed as Smalltalk code. The Smalltalk stack is represented in \openSmalltalkVM as a double linked list of stack pages which are maintained by the VM. Each stack page is represented as a Smalltalk object in the simulator.

@@

In Simulation, one of the core feature is to be able to re-use the whole Smalltalk IDE, including the browser, the inspectors and the debugger to develop and debug the VM. Most new features can be develop and debug interactively, adding code to the VM at runtime, in the simulation environment, as for normal Smalltalk programs.

@@

>>>>

The VM executable is generated in a two step process. Firstly, the Slang-to-C compiler translates the Slang code to C code, generating a few files. This first step takes several seconds. Secondly, the C compiler \footnote{depending on the platform, LLVM, GCC, MSVC} translates the C code into an executable. This second step may take a few minutes the first time. But, depending on what the programmer changed and what the C compiler has cached, it usually only takes a few seconds to recompile the C code.

The VM can actually be configured in two main flavours, interpreter-only or interpreter+JIT. Although the version with the JIT is most often used in production, the interpreter version is convenient for development purposes \footnote{and on devices that outlaw JITs}. For example, debugging the garbage collector or evaluating new language features can be done in the interpreter-only VM, avoiding JIT complexity.

@@

\paragraph{Start-up from snapshots.} 
Smalltalk is an object \emph{system}, rather than a language. The entire system, including its development tools and application code is stored in a snapshot file, which is essentially a memory dump of the entire heap. A Smalltalk system is started by loading a snapshot file into a VM. The snapshot includes objects such as the classes, the compiled methods in the form of bytecodes and the running processes. At start-up, the VM restores the state of all objects in memory, swizzling pointers as required, and resumes execution in the process active when the snapshot was created.  A new snapshot can be made during or at the end of the session.

When programming with Smalltalk, the programmer usually starts from a snapshot which contains the core libraries, the development environment and the application under development. Development of the application consists essentially in writing and editing code, which effectively installs, modifies and removes classes and compiled methods from the set of live objects. Programming may be done live, as the application under development is running, for example creating or revising methods in the debugger, and continuing.  Objects may have their shape changed on the fly as instance variables are added and removed.  \footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello \cite{Monticello}}.

  Snapshots can be used to avoid long startup times for specific bugs.  A simulation can be run to a point where the bug is about to be induced, the system snapshotted, and then multiple analyses of the bug undertaken by loading the snapshot and resuming the simulation, short cutting the set up time for the simulator.  Simulation is of the order of a thousand times slower than actual execution.

\paragraph{Interpreter-only simulation.} 
As the VM can be compiled with and without the JIT, the VM can also be simulated with or without the JIT. Since executing machine code on the processor simulators is not fast, the simulated interpreter may be faster \footnote{but JIT code is so much more optimized than bytecode interpretation that it may indeed be faster, depending on the code being executed}, and startup time and memory footprint are smaller.

@@

At start-up, the VM mmaps a memory region used for the executable code generated by the JIT (optional executable section, exists only if the JIT is enabled), for the young objects and for the old objects present in the snapshot started plus a little more space for the first \emph{tenures}\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector. Later during execution, as old space grows, new memory regions are mmap'ed at higher addresses to store other old objects. Usually at very high addresses, we find the C stack. Room in the C stack is allocated at VM start-up to hold the Smalltalk stack zone. Both stacks are disjoint and managed differently, the stack zone being broken up into small stack pages \cite{StackMappingVW}.

In the simulator, the heap is stored as a large contiguous byte array. References between objects are indices inside the byte array instead of pointers. All the Slang variables, normally translated to C variables, are simulated as Smalltalk objects. They use specific wrapper classes, such as \emph{CArrayAccessor} over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs, etc.). The Slang code is executed as Smalltalk code. The Smalltalk stack zone is represented in \openSmalltalkVM as a double linked list of stack pages which are maintained by the VM. In the interpeter configuration the stack zone is a Smalltalk Array object.  In the JIT it is in the byte array, to allow access to the stack from generated machine code.

@@

\eem{this feels repetitious.  we should say this once, and once only.  where is it best said?}
In Simulation, one of the core feature is to be able to re-use the whole Smalltalk IDE, including the browser, inspectors and debugger to develop the VM. Most new features can be developed interactively, adding code to the VM at runtime, in the simulation environment, as for normal Smalltalk programming.

\paragraph{Miscellaneous Uses}

In general, the simulator provides a toolkit for manipulating and inspecting image files.
Starting up \openSmalltalkVM from source files has never been supported. However, recent work~\cite{PharoBootstrap} allows a snapshot to be created from sources, indirectly allowing the VM to be started from source files. 
The bootstrap process uses the simulator to create and manipulate the heap and then to save the resulting snapshot.

The simulator is also used to create 64-bit images from 32-bit images, a transformation that involves replacing certain instances of classes by certain others, for example, the 64-bit system supports an immediate floating point type that represents a subset of 64-bit IEEE 754 floating point numbers whose exponent fits in the same 8-bit range as 32-bit IEEE 754 floats, whereas in the 32-bit system all floats are 8 byte boxed objects.

@@

\paragraph{JIT fidelity and the interfacing of machine code to the simulation objects}
A key requirement to enable effective development of the JIT is that the code generated during simulation be as close as possible to the code generated by the production JIT.  The code is of course not bit-identical; the address space is laid out very differently in memory to the simulation's byte array and the simulator is not subject to Address Space Layout Randomisation, a feature we augment with a synthetic clock so that time advances in lock step with code execution. Hence simulations are exactly repeatable.  But we benefit from the code generated under simulation being otherwise identical to that generated in the production VM and we go to some effort to achieve this.  The technique we use, involving illegal addresses, also allows us to interface the Smalltalk objects that comprise the Simulation, the interpreter, memory manager and jit, and variables within them, with machine code. To explain this let's consider how the interpreter references a stack frame.

The interpreter has framePointer, stackPointer and instructionPointer variables used to interpret code. When machine code wants to enter the run time, for example to have the run time look up a send, or to invoke the scavenger when eden is full, it needs to record the current stack frame, writing the native stack and frame pointers into the stackPointer and framePointer variables, and then to call the relevant routine in the run time. The run time can then access the current machine code stack frame though these variables.  In the production VM, the stackPointer and framePointer variables are C variables at known addresses in Data, and routines in the runtime have been compiled from C and exit in Text.  Referencing these variables and invoking runtime routines from machine code is therefore straight-forward.  But in the simulation they are inaccessible.  First of all they are instance variables inside Smalltalk objects, objects that have strong encapsulation, accessible only through messages, and second of all they are alongside the simulation's byte array, not within it, therefore the processor simulator has no access to these variables.

The JIT maintains a set of Dictionaries (hash maps) that map integers representing addresses to closures that get or set variables in the simulation objects, or invoke runtime routines, in both cases by sending messages to the simulation objects.  As code is JITted and specific variables and runtime routines are referenced the JIT manufactures a unique illegal address to reference the variable or routine, and uses this as the key in the relevant map, creating and storing the relevant closure accessor as the value.  When the processor simulator attempts to execute an instruction containing one of these illegal addresses, which will be some variety of read, write, call, jump or return instruction \footnote{activations of JITted machine code methods return to calling interpreter frames by returning to a trampoline that transfers control to the interpreter; hence the return address of a machine code activation above an interpreter frame has this trampoline as its return address}, it will attempt to raise an exception.  Our processor simulators are invoked through a "primitive" interface, much like an FFI call, and the attempt to raise an exeption is caught by wrapper code that returns an error result from the primitve invocation; in Smalltalk terminology the primitive fails.  The primitive failure code, Smalltalk code in the method that contains the primitive invocation, then analyses the offending instruction and constructs a Smalltalk ProcessorException object.  This Smalltalk exception has a type (\#read, \#write, \#jump, \#call), the address of the memory location being read, written, jumped to or called, a program counter (pc) holding the instruction's address, a next pc, and an accessor for the register, if any, that is the target of the read or the source of the write.  The processor simulator is invoked in the context of an exception handler for ProcessorException exceptions which invokes a Smalltalk method, handleSimulationTrap:, that invokes the relevant closure.  Once the closure has completed, the processor simulator is advanced to the nextpc and simulation resumes.  In this way, machine code generated in simulation can access arbitrary simulation objects while remaining essentially identical to the generated machine code in the production VM.

\paragraph{Machine code decoration, arbitrary breakpoints and history}
Blah, blah, blah.
